Geometric Filters for Metric Spaces

### 2018-11-29 paper/draft.tex
interim

It is known that asymptotically optimal spanner can be constructed by the following greedy algorithm: 
start with an empty graph on vertices $p_i$,
sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
until it becomes a $1+\eps$-spanner.

 In our setting this is the most expensive approach, because it requires
 computing all pairwise distances first. We propose an alternative approach (or rather a family of algorithms), which we call
 \textit{blind} spanner, meaning that the algorithm that builds a spanner does not have access to $\dist(p_i, p_j)$
 before it decides to add the edge $(p_i, p_j)$.
 The algorithm also maintains lower and upper bounds
 for all pairwise distances, which are updated after a new edge is added. and the stopping criterion
 is based on these bounds, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
 for the distance between $p_i$ and $p_j$, respectively, then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
 for all $i \neq j$.

\begin{algorithmic}
\label{alg:blind_spanner}
\Function{BlindSpanner}{$P, \eps$}
    \State {$E \gets \emptyset$}
    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
    \State {$(i,j) \gets GetEdgeToAdd$}
    \State {$v \gets \dist(p_i, pj)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \State {$UpdateBounds(i, j, v)$}
    \EndWhile
\EndFunction
\end{algorithmic}

In this pseudocode we adopt the convention that a positive number divided by 0 is $\infty$
and $\infty$ is larger than any real number,
thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is desribed in the next subsection,
now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
and the update is monotonic: no upper bound can increase, no lower bound can decrease, and, after the distance
between $p_i$ and $p_j$ is calculated, both $a_{i,j}$ and  $b_{i,j}$ are set to be $\dist(p_i, p_j)$.
The most important part of this algorithm is the way of choosing the next edge to be added, and
there are several possibilities, which give very different results.

@@

 In this subsection we explain the update procedure that we use. Suppose that
 $a_{k,l}$ and $b_{k,l}$ are valid lower and upper bounds, so that
\[
     \forall 1 \leq k < l \leq n \quad a_{k,l} \leq \dist(p_k, p_l) \leq b_{k,l}.
\]
Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$. 
To update the upper bound of some entry $b_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge. Hence, we update
\[
    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
\]
Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
time per entry.

@@

We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.
%Let $e$ denote the number of edges of a spanner on $n$ vertices;
%in the plots we visualize the ratios $e / { n \choose 2}$ and $e / n$.
%We refer to the first ratio as \textit{density} and to the second one
%as \textit{growth}. Note that

@@

The algorithm that we analyze can be roughly described as follows. 
Pick a random point $p_i$,
compute $\dist(p_i, q)$ and use this distance combined with pairwise distances $\dist(p_i, p_j)$
to update lower bounds for other distances $\dist(p_j, q)$.
If we see that the lower bound for $\dist(p_j, q)$ is larger than $\dist(p_i, q) / (1+\eps)$, then 
we can discard the point $p_j$. If all points were discarded, we found an approximate nearest neighbor;
otherwise pick a random point $p_j$ from the remaining ones and compute $\dist(p_j, q)$.
There are two possibilities: if $\dist(p_j, q) \leq \dist(p_i, q) / (1+\eps)$,
then we choose $p_j$ as the current candidate for an approximate nearest neighbor,
otherwise we keep $p_i$ in this role. In any case, the new computed distance
may also allow us to discard more than one point, as figure \ref{fig:discarded_balls} shows. In
this figure points inside the shaded smaller ball cannot be a better candidate,
and the same is true for the points in the shaded area outside of the larger ball.
The precise radii of these balls are given in the following lemma.

>>>>

In this section $P = \Set{p_1, \dots, p_n}$ is a given set of points in a metric space;
our goal is to construct a $(1+\eps)$-spanner on $P$. All algorithms that we consider 
build a spanner by starting with an empty graph on vertices $P$, to which they iteratively
add edges. We shall denote this graph (at any stage of the algorithm) by $G$.

It is known that an asymptotically optimal spanner can be constructed by the following greedy algorithm: 
sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to $G$
until it becomes a $1+\eps$-spanner. This algorithm in its most naive form is expensive in terms of $n$,
because it requires computing many shortest paths, and it is also the most expensive approach
in our setting, because it requires
 computing all pairwise distances first. We propose an alternative algorithm (or rather a family of algorithms), 
 which we call
 \textit{blind} spanner, meaning that the algorithm that builds a spanner does not have access to $\dist(p_i, p_j)$
 before it decides to add the edge $(p_i, p_j)$ to $G$.
 The algorithm also maintains lower and upper bounds
 for all pairwise distances, which are updated after a new edge is added, and the stopping criterion
 is based on these bounds, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
 for the distance between $p_i$ and $p_j$, respectively, then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
 for all $i \neq j$.

\begin{algorithmic}
\label{alg:blind_spanner}
\Function{BlindSpanner}{$P, \eps$}
    \State {$G \gets (P, \emptyset)$}
    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
        \State {$(i,j) \gets$ \Call {GetEdgeToAdd}{}
        \State {$v \gets \dist(p_i, pj)$}
        \State {Add weighted edge $(p_i, p_j, v)$ to $G$}
        \State {$UpdateBounds(i, j, v)$}
    \EndWhile
    \State \Return {$G$}
\EndFunction
\end{algorithmic}

In this pseudocode we adopt the convention that a positive number divided by 0 is $\infty$
and $\infty$ is larger than any real number,
thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is described in the next subsection,
now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
and the update is monotonic: no upper bound can increase, no lower bound can decrease, and, after the distance
between $p_i$ and $p_j$ is calculated, both $a_{i,j}$ and  $b_{i,j}$ are set to be $\dist(p_i, p_j)$.
Combined with the natural requirement that the function \textsc{GetEdgeToAdd} never
returns the same edge twice, these properties ensure that the algorithm terminates.
The most important part of this algorithm is the way of choosing the next edge to be added, and
there are several possibilities, which give very different results.

@@

 In this subsection we explain the update procedure that we use. Suppose that
 $a_{k,l}$ and $b_{k,l}$ are valid lower and upper bounds, so that
\[
     \forall 1 \leq k < l \leq n \quad a_{k,l} \leq \dist(p_k, p_l) \leq b_{k,l}.
\]
Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$. 
Now observe that for a pair $(k, \ell)$ there are three possibilities:
either the shortest path from $p_k$ to $p_\ell$ in $G$
To update the upper bound of some entry $b_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge.

Hence, we update
\[
    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
\]
Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
time.

@@

We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

@@

The algorithm that we analyze can be roughly described as follows. 
Pick a random point $p_i$,
compute $\dist(p_i, q)$ and use this distance combined with pairwise distances $\dist(p_i, p_j)$
to update lower bounds for other distances $\dist(p_j, q)$.
If we see that the lower bound for $\dist(p_j, q)$ is larger than $\dist(p_i, q) / (1+\eps)$, then 
we can discard the point $p_j$. If all points were discarded, we found an approximate nearest neighbor;
otherwise pick a random point $p_j$ from the remaining ones and compute $\dist(p_j, q)$.
There are two possibilities: if $\dist(p_j, q) < \dist(p_i, q) / (1+\eps)$,
then we choose $p_j$ as the current candidate for an approximate nearest neighbor,
otherwise we keep $p_i$ in this role. In any case, the new computed distance
may also allow us to discard more than one point, as figure \ref{fig:discarded_balls} shows. In
this figure points inside the shaded smaller ball cannot be a better candidate,
and the same is true for the points in the shaded area outside of the larger ball.
The precise radii of these balls are given in the following lemma.
